---
title: "In-class Exercises Week 1"
author: "Your "
output:
 html_document:
   theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set 1, Exercise 1

Write a set of conditional(s) that satisfies the following requirements:

* If `x` is greater than 3 and `y` is less than or equal to 3 then print "Hello world!"

* Otherwise if `x` is greater than 3 print "!dlrow olleH"

* If `x` is less than or equal to 3 then print "Something else ..."

* Stop execution if x is odd and y is even and report an error, don't print any of the text strings above.

Test out your code with

1. `x=1` and `y=2`
2. `x=1` and `y=3` 
3. `x=3` and `y=3`

**Note that because of the `stop` command, the `.Rmd` file will not knit when errors occur unless you set the chunk option `error=TRUE`**

```{r Set1Exe1, error=TRUE}
# write your code for problem set 1 exercise 1 here
x <- 3
y <- 3
if ((x%%2) == 1 & (y%%2) == 0){
  stop("Error Message.")
}else{
  if(x > 3){
    if(y <= 3){
      print("Hello World!")
    }else{
      print("!dlroW olleH")
    }
  }else{
    print("Something else. . .")
  }
}
```

## Set 1, Exercise 2

Write a set of conditional(s) in `R` to solve equations of the form $$a x^2 + b x + c=0,$$ which accounts for possible errors that might occurr in the calculation (you may use `if`, `else if`, `else`, `stop`, `stopifnot`)

Test out your code with

1. `a=1`, `b=0` and `c=1`
2. `a=1`, `b=0` and `c=-1`
3. `a=9`, `b=4` and `c=1`
4. `a=0`, `b=3` and `c=3`
4. `a=1`, `b=-4` and `c=4`


**Again, because of the `stop` command the `.Rmd` file will not knit when errors occur unless you set the chunk option `error=TRUE`**

```{r Set1Exe2, error=TRUE}
# write your code for problem set 1 exercise 2 here
a = 1
b = 4
c = 1

testvar <- (b^2 - 4*a*c)
roots <- NULL
if (testvar<0){
  stop("no real roots")
}else{
  if(a==0){
    roots <- -c/b
    cat("equation is linear (i.e., a = 0 )","\n")
  }else{
    roots <- (-b+sqrt(testvar)*c(-1,1))/(2*a)
  }
  cat("the solution is: x=", paste(unique(roots),sep="",collapse=" and "),"\n")
}

```

# Set 2

Below is a vector containing all prime numbers between 2 and 100:


```{r}
primes = c( 2,  3,  5,  7, 11, 13, 17, 19, 23, 29, 31, 37, 41, 
      43, 47, 53, 59, 61, 67, 71, 73, 79, 83, 89, 97)
```


If you were given the vector 

`x = c(3,4,12,19,23,51,61,63,78)`, 
</div>
write the R code necessary to print only the values of `x` that are *not* prime (without using subsetting or the `%in%` operator). 

Your code should use *nested* loops to iterate through the vector of primes and `x`.


```{r Set2, echo=T}
# write your code for problem set 2 here

primes = c(2, 3, 5, 7, 11, 13, 17, 19, 23, 29, 31, 37, 41, 43, 47, 53, 59, 61, 67, 71, 73, 79, 83, 89, 97)
x = c(3, 4, 12, 19, 23, 51, 61, 63, 78)

res <- c()
j <- 1
b <- x[1]

for(a in primes){
  if(j>length(x)){
    break
  }
  while( (b<=a) & (j<= length(x)) ){
    if(b!=a){
      res <- c(res,b)
    }
    j <- j+1
    b <- x[j]
  }
}
res
```

