---
title: "week4_dontknockmydoor"
author: "Randi Bolt"
date: "4/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Plan for today
 * Visualization in base R 
 
 ## Data Visulaization in R before ggplot2
 * R is a software for statistics it comes with extensive plotting capabilities
 * MANY plotting functions are installed in the graphics package, whcih ships with base R
 * Look into the help files for the functions in this package using 
```{r}
library(help = "graphics")
```
 
* Here are a few examples of visualizations functions in the graphics package: 
plot, lines, points, abline, boxplot, pairs, matplot, barplot, pie, rasterImage, coplot, cdplot, mosaicplot, polygon

## Data Visualization before ggplot2
plot
* plot is gereric function for plotting R objects
* Functions in R can be designed in such a way that the same function can have a completely different behavior depending on the object it is used with
* Here is a lit with some of the objects it can be applied to 
```{r}
methods(plot)
```

## Default
* scatterplots
* structure : plot( x, y, z)
+ x is a variable in the x axis 
+ y in the variable 
+ ... represents other graphical parameters (see ?par for an extensive list)
```{r}
?par
```

## Data visualization before ggplot2
Eample: the cars dataset
* lets load the build- in data **cars** which loads as a dataframe, a type of object mention earlier. Then, we can look at it in a couple different ways. 
* data(cars) loads this dataframse into the **Global Enviroment** as a promise. Promises are unevaluated arguments. 
```{r}
data(cars)
str(cars, 5) # telling the structure 
```
```{r}
head(cars,4) #shows first four rows
summary(cars) # summary stats for each var (speed and distance)
```
## plot options
```{r}
plot(x = cars$speed, y = cars$distance,
     xlab = "Speed (mph)", # x label 
     ylab = "Stopping Distance (ft)", # y label 
     main = "Speeds and Stopping distance of cars",
     type = "p", # type of plot i.e. points 
     lty = 1, lwd = 1, # lty is the line type, 1 is a solid line, 2 dot, 3 dash
                       #lwd is line width 
     pch = 16, #plotting character : i.e. circle, squares, triangles, stars
     cex = 1, cex.axis = 1, cex.lab = 1, #character expansion is the point size
               # can highlight specific points(make them bigger)
               # character expansion for axis : want axis to be bigger
               # cex. for labels 
     col = "firebrick") #color
```
## Modify 
Plot type
* 'p' = points
* 'l' = lines
* 'b' = both 

## plot option: cex

```{r}
# as speed increases size increases 
obj.size <- seq(0.5,5,length.out = length(cars$speed)) 
plot(x = cars$speed, y = cars$distance,
     xlab = "Speed (mph)",  
     ylab = "Stopping Distance (ft)", 
     main = "Speeds and Stopping distance of cars",
     type = "p",
     lty = 1, lwd = 1,
     pch = 16, 
     cex = obj.size, cex.axis = 1, cex.lab = 1, 
     col = "firebrick") 
```
```{r}
# farther the distance the bigger the plot 
obj.size <- log(cars$dist)/4
plot(x = cars$speed, y = cars$distance,
     xlab = "Speed (mph)",  
     ylab = "Stopping Distance (ft)", 
     main = "Speeds and Stopping distance of cars",
     type = "p",
     lty = 1, lwd = 1,
     pch = 16, 
     cex = obj.size, cex.axis = 1, cex.lab = 1, 
     col = "firebrick") 
```
##plot options : color
```{r}
col.vec <- rep(c("firebrick","forestgreen","cornflowerblue"),
               times=c(sum(cars$speed<10),
                       sum(cars$speed>=10&cars$speed<17),
                       sum(cars$speed>=17)))
plot(x = cars$speed, y = cars$distance,
     xlab = "Speed (mph)",  
     ylab = "Stopping Distance (ft)", 
     main = "Speeds and Stopping distance of cars",
     type = "p",
     lty = 1, lwd = 1,
     pch = 16, 
     cex = 1, cex.axis = 1, cex.lab = 1, 
     col = col.vec) 
```
## Adding a legend
```{r}
plot(x = cars$speed, y = cars$distance,
     xlab = "Speed (mph)",  
     ylab = "Stopping Distance (ft)", 
     main = "Speeds and Stopping distance of cars",
     type = "p",
     lty = 1, lwd = 1,
     pch = 16, 
     cex = 1, cex.axis = 1, cex.lab = 1, 
     col = col.vec)
legend("topleft", #location, topright, bottom, bottomright (look at help)
                  #can plot as vector as well within 
       bty = "n", #n mean no box (daniel thinks box looks clunky)
       pch=c(16,16,16),#point type 
       col=c("firebrick","forestgreen","cornflowerblue"),
       legend=c("speed<10","10<=speed<17", "speed>=17"))
```
## boxplot
```{r}
par(mfrow=c(1,2),mai=c(1,0.5,0.1,0.1)) # parameters
                      # modify window showing plot
                      # mai is internal margins 
                      # mfrow : graphical device of 1 row with two colomns 
boxplot(decrease ~ treatment, data = OrchardSprays, col = "cornflowerblue",
        log = "y",cex.axis=0.7,cex.lab=0.7,notch=F)
## horizontal=TRUE, switching  y <--> x :
boxplot(decrease ~ treatment, data = OrchardSprays, col = 1:8,
        log = "x", horizontal=TRUE,cex.axis=0.7,cex.lab=0.7,notch=F)
```
### Other plots: `curve`

```{r curvefn, echo=T, fig.height=2.5, fig.width=7.5}
par(mfrow=c(1,3),mai=c(0.9,0.4,0.1,0.1))
curve(expr=sin, from=-2*pi, to=2*pi, xname = "t",cex.axis=0.7, cex.lab=0.7)
curve(expr=tan, xname = "t", from=-2*pi, to=2*pi, cex.axis=0.7, cex.lab=0.7)
myfn <- function(t){tan(t)*sin(t)}
curve(expr=myfn, xname = "t", from=-2*pi, to=2*pi, cex.axis=0.7, cex.lab=0.7)
```
### Other plots: `hist`
```{r histfn, echo=T, fig.height=3, fig.width=6}
par(mfrow=c(1,2),mai=c(1,0.5,0.1,0.1))
x <- rchisq(1000, df = 4)
hist(x, freq = FALSE, ylim = c(0, 0.2),col="orange",main="")
#hist and curve combined
hist(x, freq = FALSE, ylim = c(0, 0.2),col="orange",main="")
lines(density(x,from=0, to=20), col = "blue3", lty = 1, lwd = 3)

# density : dont caluclulate values of the tail 
# ylim : limit of the y axis 
```
### Other plots: `pairs`

```{r pairsfn, echo=T, fig.align='center',fig.height=4, fig.width=4}
par(mai=c(0.1,0.1,0.1,0.1))
pairs(iris[1:3], cex=0.5, cex.labels = 1,cex.axis=0.7,
      pch = 21, bg = c("red", "green3", "blue")[unclass(iris$Species)])
```

```{r}
#real type of a vector is an integer vector 
data(iris)
iris$Species
unclass(iris$Species)
```
### Other plots: `dotchart`
<div class="columns-2">
```{r}
VADeaths
```

```{r, eval=F, echo=T}
par(mai=c(0.4,0.1,0.4,0.1))
dotchart(VADeaths, bg = "skyblue",
         cex=0.7, cex.axis=0.1,
         main = "Death Rate VA - 1940")
```
<p class="forceBreak"></p>

```{r dotchartfn, eval=T, echo=F, fig.align='center',fig.height=3.5, fig.width=3.5}
par(mai=c(0.4,0.1,0.4,0.1))
dotchart(VADeaths, bg = "skyblue",
         cex=0.7, cex.axis=0.1,
         main = "Death Rate VA - 1940")
```
</div>
### Other plots: `matplot`

```{r matplotfn, echo=T, fig.height=3, fig.width=6}
#throw in matrix and give one function per line of matrix 
par(mfrow=c(1,2),mai=c(0.4,0.4,0.1,0.1))
#mfrow : enter figures by row (can do mfcol as well)
#takes two vectors, and combine every combination of first and second vector 
sines <- outer(1:20, 1:4, function(x, y) sin(x / 20 * pi * y))
# matrix with 20 rows and four columns
# fills in from left to right
matplot(sines, pch = 1:4, type = "o", col = rainbow(ncol(sines)),
        cex=0.5, cex.axis=0.7)
        #rainbow puts as many colors as sines in matrix (which is four)
        #type 'o' : is for overlay , could also have something else here
matplot(sines, type = "b", pch = 21:23, col = 2:5, bg = 2:5, 
        cex=0.5, cex.axis=0.7, main = "")
        #pch is plotting character 
```
## rainbow
```{r}
rainbow(6)
```

```{r}
#sines 
outer(1:20, 1:4, function(x, y) sin(x / 20 * pi * y))
```

### Other plots: `barplot`

```{r barplotfn, echo=T, fig.height=3, fig.width=6}
par(mfrow=c(1,2),mai=c(0.9,0.4,0.1,0.1))
barplot(GNP ~ Year, data = longley, cex=0.5, cex.axis=0.7, cex.lab=0.7)
        #formual rep. , give me bar plot by year that shows longly data
        # before ~ represents y (GNP) and after represents x (YEAR)
barplot(cbind(Employed, Unemployed) ~ Year, data = longley, cex = 0.5,
        cex.axis=0.7,cex.lab=0.7)
        #plotting two different response variables 
        #(employed and unemployed) by year
```
### Other plots: `mosaic`

```{r mosaicfn, echo=T, fig.height=4, fig.width=4}
#good for category data (only a few)
par(mfrow=c(1,1),mai=c(0.9,0.4,1,0.4))
mosaicplot(~ Sex + Age + Survived, data = Titanic, main="",color = TRUE)
```
### <span style="color:cornflowerblue">Figures with multiple panels:</span> `mfrow` (or `mfcol`)
```{r, eval=F,echo=T, fig.height=2.5, fig.width=7.5}
par(mfrow=c(1,3)) #dim of layout matrix 1 row and 3 columns, fill by row not col
plot(x = cars$speed, y = cars$dist,xlab = "", ylab = "", main = "",
     type = "p", pch = 16, col = "firebrick") 
plot(x = cars$speed, y = cars$dist,xlab = "", ylab = "", main = "",
     type = "p", pch = 16, col = "firebrick") 
abline(reg=lm(dist~speed,data=cars),col="forestgreen")
      # overlaying regression line of slope ontop of previous plot 
plot(x = cars$speed, y = cars$dist,xlab = "", ylab = "", main = "",
     type = "p", pch = 16, col = "firebrick") 
lines(lowess(cars),col="cornflowerblue")
```
### <span style="color:cornflowerblue">Figures with multiple panels:</span> `mfrow` (or `mfcol`)
```{r, eval=T,echo=F, fig.height=2.8, fig.width=8.4}
par(mfrow=c(1,3))
plot(x = cars$speed, y = cars$dist,xlab = "", ylab = "", main = "",
     type = "p", pch = 16, col = "firebrick") 
plot(x = cars$speed, y = cars$dist,xlab = "", ylab = "", main = "",
     type = "p", pch = 16, col = "firebrick") 
abline(reg=lm(dist~speed,data=cars),col="forestgreen",lwd=2)
plot(x = cars$speed, y = cars$dist,xlab = "", ylab = "", main = "",
     type = "p", pch = 16, col = "firebrick") 
lines(lowess(cars),col="cornflowerblue",lwd=2)
```
### <span style="color:cornflowerblue">Figures with multiple panels:</span> the `layout` function

```{r, eval=FALSE}
layout(mat, 
       widths = rep.int(1, ncol(mat)),
       heights = rep.int(1, nrow(mat)), 
       respect = FALSE)
```
### <span style="color:cornflowerblue">Figures with multiple panels:</span> the `layout` function

```{r, echo=T,eval=T,fig.height=4,fig.width=4}
nf <- layout(matrix(c(2,0,1,3),2,2,byrow = TRUE), widths=c(3,1), heights=c(1,3), TRUE)
            # matrix is saying put two upper left, nothing in upper right
            # 1 in lower left, and three in lower right 
            #widths / heights : for object 
layout.show(nf)
```
### <span style="color:cornflowerblue">Figures with multiple panels:</span> the `layout` function

```{r, echo=T,eval=F, fig.height=4.5,fig.width=4.5}
nf <- layout(matrix(c(2,0,1,3),ncol=2,nrow=2,byrow = TRUE), 
             widths=c(3,1), heights=c(1,3), 
             respect=TRUE)

#sample values from two normal distributions
x <- pmin(3, pmax(-3, stats::rnorm(50))) # 50 standard normal and then put 
y <- pmin(3, pmax(-3, stats::rnorm(50))) # b/t -3 and 3 for plotting looks good

#generate and store histograms for x and y
xhist <- hist(x, breaks = seq(-3,3,0.5), plot = FALSE) #storing value of hist
yhist <- hist(y, breaks = seq(-3,3,0.5), plot = FALSE)
top <- max(c(xhist$counts, yhist$counts)) #used to defin. upper limit for hists

xrange <- yrange <- c(-3, 3)
    # initializing ranges to use later
par(mar = c(3,3,1,1)) #mar : margins (unsure of order)
plot(x, y, xlim = xrange, ylim = yrange, xlab = "", ylab = "")
    #scatter plot in the middle 
par(mar = c(0,3,1,1))
barplot(xhist$counts, axes = FALSE, ylim = c(0, top), space = 0)
    # azes = FALSE : dont plot the axis 
    # ylim : 0 to top
    #space = 0 : no spaces b/t bars
par(mar = c(3,0,1,1))
barplot(yhist$counts, axes = FALSE, xlim = c(0, top), space = 0, horiz = TRUE)
```

## what xhist is storing 
```{r}
x <- pmin(3, pmax(-3, stats::rnorm(50))) # 50 standard normal and then put 
y <- pmin(3, pmax(-3, stats::rnorm(50)))
xhist <- hist(x, breaks = seq(-3,3,0.5), plot = FALSE)
xhist
```

