---
title: "Week2_ElectricBoogaloo"
subtitle: "Week 2 Notes"
author: "Randi Bolt"
output: 
  html_document: 
  theme: readable
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Atotmic Vectors can be constructed using concatenate c()
```{r}
c(1,2,3)
c("Hello", "World")
#nested concatenates
c("a",c("b",c("c")))
```

## Testing types
typeof(x) 
mode(x) 
```{r}
typeof(TRUE)
mode(TRUE)
typeof("hello")
mode("hello")
typeof(1.33)
mode(1.33)
typeof(7)
mode(7)
typeof(7L)
mode(7L)
```

## Type Coercion
character <- double <- int <- boolean
```{r}
typeof(c(1,"Hello"))
typeof(c(FALSE,3L))
typeof(c(1.2, 3L))
```

## Operator Coercion
```{r}
3.1+3L
log(TRUE)
TRUE & 7 # true bc/ not 0
FALSE | 5 #flase or false is true
```

## Explicit Coercion
as. 
```{r}
as.logical(5.2)
as.character(TRUE)
as.integer(pi)
as.numeric(FALSE)
as.double("7.2")
as.double("one")

```

## Missing Values 
NA : used to represent missing values. By default NA is logical. 
```{r}
typeof(NA)
typeof(c("a",NA))
typeof(NA+1)
typeof(NA_character_)
typeof(NA_real_) #never see type appended
typeof(NA_integer_)
```

## Stickiness of Missing Values
```{r}
1+NA
1 / NA
NA*5
sqrt(NA)
mean(c(1,2,3,NA))
3^NA
```

## Conditionals and Missing values
NA's can cause control flow issues
```{r, error=TRUE}
if(2!=NA){"Here"}
if(all(c(1,2,NA,4) >= 1)){"There"}
if(any(c(1,2,NA,4) >= 1)){"There"}
```

## Testing for missing valuse
is.na
```{r}
is.na(NA)
is.na(1)
is.na(c(1,2,3,NA))
any(is.na(c(1,2,3,NA)))
all(is.na(c(1,2,3,NA)))
```

## Other special (double) values
NaN - not a number, Inf - positive infinity, -Inf - negative infinity
```{r}
pi/0
0/0
1/0+1/0
1/0-1/0
NaN + NA
```

## Testing for inf and NaN
```{r}
is.finite(NA)
is.finite(1/0+1/0)
is.finite(1/0-1/0)
is.nan(1/0-1/0)
```

## Coercion for Infinity and NaN
```{r}
as.integer(Inf)
as.integer(NaN)
as.logical(Inf)
as.logical(-Inf)
as.logical(NaN)
as.character(Inf)
as.character(NaN)
```

## Exercise Set 2 Proble 2

```{r}
paste("label ", 1:30 , sep= "", ",")
paste("fn", 1:30, sep= "")
```

## Generic Vectors
Lists : are generic 1 dimensional 
```{r}
list("A", c(TRUE,FALSE), (1:4)/2, function(x) x^2)
```

## Structure

```{r}
str(list("A", c(TRUE,FALSE), (1:4)/2, function(x) x^2))
```

## Recursive Lists
```{r}
str(list(1,list(2, list(3,4), 5)))
```

## List Coercion
```{r}
str(c(1, list(4, list(6,7))))
unlist(c(1, list(4, list(6,7))))
unlist(list(1, list(2, list(3, "Hello"))))
```

## Named Lists

```{r}
str(list(A=1, B = list(C = 2, D = 3)))
list("knock knock" = "whos's there?")
#names gives names of list in highes heirarchy
list(ABC = 1, DEF = list(H=2, I=3))
names(list(ABC = 1, DEF = list(H=2, I=3)))
```

# Exercise Set #2 Problem #3
```{r}
str(list(firstName = "John", 
         lastName = "Smith", 
         age = 25L, 
         address = list(streetAddress = "31 3rd Street", 
                        city = "New York", 
                        state = "NY", 
                        postalCode = 10021L), 
         phoneNumber = list(type = "home", 
                            number = "212 555 5555"), 
                       list(type = "fax",
                             number = "646 555 5555")))
```

## Objects with Attributes
e.g. class, comment, dim, dimnames, names

```{r}
(x <- c(L=1, M=2, N=3))
attr(x, "names") <- c("A", "B", "C")
names(x)
```

## Attributes
```{r}
str(x)
attributes(x)
str(attributes(x))
```

## Factors
* class = "factor"
* levels = a character vector
```{r}
#highest degree a person's obtained 
(x <- factor(c("BS", "MS", "PhD", "MS")))
#int with levels attributes. Each level assigned a level
#factor is a function. 
str(x)
#assign levels
(x <- factor(c("BS", "MS", "PhD", "MS"), levels=c("PhD", "MS", "BS")))
attributes(x)
```

## Exercise 2 Problem 4
```{r}
x <- c(2L, 4L, 1L, 1L, 1L, 1L, 2L, 1L, 2L, 4L, 4L, 3L, 3L, 3L )
attributes(x) <- list(class="factor", 
                      levels=c("Sunny",
                              "Mostly Sunny",
                              "Mostly Cloudy",
                              "AM Showers"))
str(x)
x
```

## Matrices
Vectors with dimensional attribute (2D version of atomic vectors). Can only take objects of the same type. 
```{r}
(m<- matrix(c(1,2,3,4), ncol=2, nrow=2))
```

```{r}
attributes(m)
```

## Column major ordering
```{r}
#takes values by colomn
colmaj <- matrix(c(1,2,3,4), ncol=2, nrow=2)
colmaj
c(colmaj)
#takes values by row
rowmaj <- matrix(c(1,2,3,4), ncol=2, nrow=2, byrow=TRUE)
rowmaj
c(rowmaj)
```

## Data Frames
* R handles heterogeneous tabular data using data frames
* A data frame in R is a list of equal length vectors
* Data frames are generated when reading a file into R, but can also be built from scratch
```{r}
df <- data.frame(x=1:3, y=c("a", "b", "c"))
str(df)
typeof(df)
attributes(df)
```

## Cooking up your own data.frame
```{r}
df2 <- list(x=1:3, y =factor(c("a","b","c")))
attr(df2,"class") <- "data.frame"
df2
str(df2)
attributes(df2)
class(df2)
# if attributes aren't there then its unsure where to put things
attr(df2, "row.names") <- 1:3
df2
str(df2)
attributes(df2)
```

## In class example
```{r}
df3 <- list(x = 1:4, y = c("a", "b", "c", "d"))
attr(df3,"class") <- "data.frame"
attr(df3,"row.names") = 1:4
df3
str(df3)
```

## Characters vs Factors
```{r}
df <- data.frame(x=1:3, y=c("a", "b","c"),stringsAsFactors = FALSE)
str(df)
```

## Length Coercion 
If the longer length is a multiple of a the shorter then this coercion will occur without any warnings / errors. 
```{r}
x <- 1:3
y <- 2:3
x+y
```

## Length Coercion 
```{r}
data.frame(x=1:3, y=c("a"))
# bad : data.frame(x=1:3, y=c("a","b"))
```

Growing data frames
Add rows or columns using rbind and cbind
```{r}
df <- data.frame(x=1:3, y=c("a","b","c"))
df
rbind(df, c(TRUE,FALSE))
df
#true c. to 1, and false c. to character "FALSE"
#have to be careful, if you convert into a character and try to do math it wont work 
cbind(df,z=TRUE)
df
```

## Growing data frames
```{r}
df1 <- data.frame(x = 1:3, y = c("a", "b", "c"))
df2 <- data.frame(m = 3:1, n = c(TRUE, TRUE, FALSE))
cbind(df1,df2)
```
## What will happen? 
* remember columns need to have the same name attribute 
```{r}
df1 <- data.frame(x = 1:3, y = c("a","b","c"))
df2 <- data.frame(m=3, n= TRUE)
cbind(df1,df2)
```

