---
title: "Week 3 Notes : Part 2"
author: "Randi Bolt"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Frame Subsetting: 
If provided with a single value, data frames assume you want to subset a column or columns - multiple values then the data frame is treated as a matrix. 

```{r}
df <- data.frame(a= 1:2, b = 3:4)
df[1]
df[[1]]
df[,"a"]
df[2]
df["a"]
df[1,]
typeof(df[1,]) # list 
typeof(df[,1]) # int
#behave more like a matrix : otherwise it will simplify to vector or single value 
df[ ,"a", drop=FALSE]
df[c("a","b","a","a","a")]
```
# Preserving vs Simplifying Subsets
```{r}
# atomic vector 
x[[1]] #simplifying (only pulling a single value, losing name)
x[1] #preserving, and can pull more than one thing
#List
x[[1]] #simplifying
x[1] #preserving
#Matrix / Array
x[1,] #simplifying
x[,1]#preserving

```

# In class example
```{r}
x <- matrix(1:15, nrow=5)
x
x[1:3,]

x[1, drop=FALSE]
```

# Subsetting and assignment 
```{r}
x <- c(1,4,7)
x[2]<- 2
x
x[x %% 2 !0] = x[x %% 2 != 0] +1
x[c(1,1)] = c(2,3)
```

# Class Example
```{r}
x <- c(1,4,7)
```
# Subsetting and assignment Cont. 
```{r}
x <- 1:6
x[c(2,NA)] <- 1
x
```
Resetting vector, and then subset vector removing position 1 and 3 and assigning to that subsector 3 
```{r}
x <- 1:6
#minus is compliment, give me not one and four
x[c(-1,-4)]<-3
x
```
Recycling position one is true and then position 2 is NA for 6 positions. 
```{r}
x <- 1:6
x[c(TRUE,NA)] <- 1 
```
Repopulating the entries 6 to 1 instead of 1 to 6 
```{r}
x <- 1:6
x[] <- 6:1
x
```

# Deleting list (df) elements
```{r}
#b recycles value true and fills vector 
df <- data.frame(a=1:2, b=TRUE, c=c("A","B"))
df
#[[]] modifying object internally 
df[["b"]] <- NULL
df
df[,"c"] <- NULL
df
```
# Subsets and Subsets
```{r}
df <- data.frame(a = c(5,1,NA,3))
#finding values = to five and replace with zero
df$a[df$a == 5] <- 0
df
```

another way to do this
```{r}
df[1][df[1] == 3]<- 0
df
```
# Permutation test for two group means
In an experiment on memory students were given lists of 24 words to memorize. After hearing the words they were assigned at random to different groups:
* One group of 12 students took a nap for 1.5 hours while 
* A second group of 12 students stayed awake and was given a caffeine pill. 
These data contain the number of words each perticipant was able to recall after the break. 
```{r}
Memory <- read.csv("http://www.mosaic-web.org/go/datasets/SleepCaffeine.csv")
summary(Memory)
```

```{r}
get_streak <- function(x){
  y <- rep(0,length(x))
  y[x == "H"] <- 1
  y <- c(0, y, 0)
  wz <- which(y == 0)
  streak <- diff(wz) - 1
  return(streak)}

kobe_streak <- get_streak(kobe$basket)
kobe_streak
barplot(table(kobe_streak))
```

## Another percentage lol 
```{r}
hits <- 0
for (x in 1:length(kobe$basket)){
    if(kobe$basket[x]=="H"){
        hits <- hits +1 
        next
    }
}
hit_percent = (hits / length(kobe$basket))*100
hit_percent
```

