---
title: "Week3_CodeeMagee"
subtitle : "Week 3 Notes : More Subsetting and Intro to Data Vis"
author: "Randi Bolt"
date: "4/13/2021"
output: 
  html_document:
  theme: readable 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Subsetting in General

* R has different subsetting operators([, [[, and $). The behavior of these operators will depend on the object they are being used with. 

In general there are 6 different data types
* pos. int., neg. int. , logical values, epmty/ NULL, Zero, character values " names" 

### Positive Integer subsetting
```{r}
x <- c(1,4,7)
x[c(1,3)]
x[c(1,1)]
x[c(1.9,2.1)]
y <- list(1,4,7)
str(y[c(1,3)])
str(y[c(1,1)])
str(y[c(1.9,2.1)])
```
### Negative Integer subsetting
```{r}
x <- c(1,4,7)
x[-1]
x[-c(1,1)]
x[c(-1,-1)]
y <- list(1,4,7)
str(y[-1])
str(y[-c(1,3)])
# Error: x[c(-1,2)]
```

### Logical Value subsetting
```{r}
x <- c(1, 4, 7, 12)
x[c(TRUE,TRUE,FALSE,TRUE)]
x[c(TRUE,FALSE)]
#test if even, and print even values
x[x %% 2 == 0]
y <- list(1,4,7,12)
str(y[c(TRUE, TRUE, FALSE, TRUE)])
str( y[c(TRUE,FALSE)])
# Error : mod doesn't work with list 
# str( y[y %% 2 == 0])
```
 ## In class example for cahracter strings
 
```{r}
x <- c("a", "b", "c")
x[x=="a"]
x[x>"a"]
```
 ### Empty subsetting 
```{r}
x <- c(1,4,7)
x[]
# Put one in every entry of vector 
x[] <- 1
x
y <- list(1,4,7)
str(y[])
```
 
### Zero Subsetting
```{r}
x <- c(1,4,7)
x[c(0,1)]
y <- list(1,4,7)
y[c(0,2)]
```

### Character subsetting
```{r}
x <- c(a=1, b=2,c=7)
x["a"]
x[c("b","c")]
y <- list(a=1,b=4,c=7)
str(y[c("a","a")])
str(y[c("b","c")])
```

### Out of bound subsetting
```{r}
x <- c(1,4,7)
x["a"]
x[c(1,4)]
y <- list(1,4,7)
str(y[4])
str(y["a"])
str(y[c(1,4)])
```

### Missing and NULL subsetting
```{r}
x[NA]
x[NULL]
x[c(1,NA)]
str(y[NA])
str(y[NULL])
str(y[c(1,NA)])
```

### [ vs. [[ with Atomic Vectors
```{r}
x <- c(a=1, b=4, c=7)
#enter position in form 
x[[1]]
#value and its name , keep structure 
x[1]
#only access value
x[["a"]]
# error : [[ only points to one particular element
# x[[1:2]]
```

### [ vs. [[ with Generic Vectors
```{r}
y[2]
y[[2]]
y[["b"]]
# Error : out of bounds 
# y[[1:2]]
y <- list(a=1, b=c(4,6,5),c=7)
y[2]
y[[2]]
y[[2:3]]
y[c(2:3)]
y$a
```

## Functions

### When to use functions

The goal of a function should be to encapsulate a small reusable piece of code.
###### Some tips
* Name should make it clear what the function does
* Functionality should be simple enough to be quickly understood. 
* The smaller and more modular the code the easier it will be to reuse elsewhere. 
* Better to change code in one location than code everywhere

### Function Parts 
The two parts of a function are the arguments (formals) and the code (body)
```{r}
gcd <- function(long1, lat1, long2, lat2){
  R <- 6371 # Earth mean radius in km
  #distance in km
  acos(sin(lat1)*sin(lat2)+cos(lat1*cos(lat2)*cos(long2-long1)))
}
formals(gcd)
body(gcd)
```

### Return values
There are two ways of returning values in R: explicit or implicit return values
```{r}
# explicit - includes one or more return statements
f<- function(x){
  return(x*x)
}
f(4)
# implicit - value of the last statment is returned 
f <- function(x){
  y <- x*x
}
# wont be able to access 
f(4)
```

### Returning multiple values
If we want a function to return more than one value we can group things using either a vector or a list
```{r}
f <- function(x){
  c(x, x^2,x^3)
}
f(2)
f(2:3) # assumes whole thing is vector 
```
### In class examples 
```{r}
f <- function(x){
  res1 <- rep(NA,length(x)*3)
  for(i in seq_along(x)){
    res1[((i-1)*3+1):(3*i)]<-c(x[i], x[i]^2, x[i]^3)
  }
  res1
}
f(2:3)
```

### Argument names
When defining a function we implicitly define names for arguments. We can use these names to pass arguments in a different order. 
* be careful if you don't specify names, and forget the order it runs through the function 
```{r}
f <- function(x,y,z){
  paste0("x=",x,"y=",y,"z=",z)
}
f(1,2,3)
f(z=1,x=2,y=3)
f(y=2,1,3)
f(y=2,1,x=3)
# error : unused m
# f(1,2,3, m=1)
```

### Argument defaults
```{r}
f <- function(x, y=1, z=1){
  paste0("x=",x,"y=",y,"z=",z)
}
# error
# f()
f(x=3)
f(y=2,2)
```

### Scope
R has generous scoping rules, if a variable isn't in the function's body, R will look for it in the next higher scope, and so on
```{r}
# f is only a function of f because y has previously been defined 
y <- 1
f <- function(x){
  x+y
}
f(3)
# y variable inside function takes precedent 
g <- function(x){
  y<-2
  x+y
}
g(3)
```
Variables defined within a scope only persist for the duration of that scope, they do not overwrite variables at higher scopes (unless you use the global assignment operator <<-, which you shouldn't)
```{r}
x <- 1
y <- 1
z <- 1
f <- function(){
  y <- 2
  g <- function(){
    z <- 3
    return(x + y + z) # applies functions
  }
  return(g()) #returns function 
}
f()
c(x,y,z)
```
#### Modified Example
```{r}
x <- 1
y <- 1
z <- 1
f <- function(){
  y <- 2
  g <- function(){
    z <- 3
    x + y + z # do not need return function 
  }
  g() #returns function 
}
f()
c(x,y,z)
```

# Break 

## Lazy Evaluation 
Arguments to R functions are lazily evaluated - meaning they are not evaluated until theyare used
```{r, error=TRUE}
f <- function(x){
  cat("Hello world!\n")
  x
}
# Error
f(stop())
```
### Getting Help 
Prefixing any function name with a ? will open the related help for that function 
```{r}
?'+'
?sum
summary.aov
```

### Less Helpful Examples
```{r}
list
'['
sum
'+'
```

### About Homework
* Your documents shold be pleasant for a peer to look at, with some organization using sections or lists, and all plots labeled clearly
* Use chunk options echo and results to limit the code/output you should in the .html file
* Discussion of specific values should be summarized in sentences in your text (use inline code) - not as printed code and output - and rounded so ans not to be absurdly precise (e.g. use round x with round(x,2)). 
* Use naming format specified 

### Peer Review: Grading Rubric 

Once the submission deadline has passed I will randomly assign graders to annonymized assignments and give you access to the files 

0 - No homework turned anything in.
1 - Turned in but low effor, poorly presented with nonfunctional code and ignoring directions.
2 - Decent effort, well presented all code works and followed directions with some minor issues.
3 - Nailed it! 

*Peer review will be next week 

### Acknowledgements
* Colin Rundel
* Jenny Bryan
* Daniel Taylor-Rodriguez
* Advanced R book
* Programming Exercises for R 

*note : example of this hw is in the r book, but do not copy 

# Week 3 : More Subsetting and Intro to Data Vis
## Subsetting
### Subsetting Matrices
```{r}
# not a type preserving operations 
(x<- matrix(1:6, nrow=2, ncol=3))
x[1,3]
x[1:3]
x[1:2,1:2]
x[,1:2]
# will convert vector into a single row, so be carful doing functions with matrices 
x[-1,-3]
```

### Preserving Subsetting 
Most of the time, R's [ subset operator is a preserving operator, in that the returned object will have the same type as the parent. Confusingly, when used with a matrix or array [ becomes a simplifying operator (does not preserve type) - this behavior is controlled by the drop argument. 
```{r}
x[1,]
# This is something you will use in practice 
x[1, , drop=TRUE]
x[1, , drop=FALSE]
str(x[1, ])
str(x[1, , drop=TRUE])
str(x[1, , drop=FALSE])
```
### In Class example
```{r}
x <- matrix(1:6, nrow=2, ncol=3)
colnames(x)<- paste("c",1:3,sep="")
rownames(x)<- paste("r",1:2,sep="")
x
```

### Factor Subsetting
```{r}
(x<- factor(c("BS","MS","PhD","MS")))
x[1:2]
# drops levels that aren't inside of new vector 
x[1:2, drop=TRUE]
```

## Next Time
Data Frame Subsetting 
